---
title: "DESCRIPTIVE AND ANALYSIS For MALARIA RISK MAPPING"
author: "Lindah"
date: "2026-04-28"
output: pdf_document
pdf_document:
    latex_engine: xelatex
header-includes:
  - \usepackage{newunicodechar}
  - \newunicodechar{−}{-}
  - \newunicodechar{–}{--}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo = TRUE)

if (capabilities("cairo")) {
  knitr::opts_chunk$set(dev = "cairo_pdf")}
```

```{r}

# Load libraries
library(tidyverse)   
library(lubridate)  
library(skimr)       
# Load the data
data <- read.csv("/home/student25/Documents/THESIS/MSC_THESIS_package_corrected - Revised/SPATIO-TEMPORAL THESIS/DATA/kenya_Malaria_panel.csv")

# Checking the  structure
str(data)
head(data)
```
\section*{CONVERTING VARIABLE CLEANING AND FORMAT VARIABLES}
```{r}

data <- data %>%
  mutate(
    county = as.factor(county),     
    date = as.Date(date),         
    year = year(date),
    month = month(date)
  )

# Confirm structure
str(data)
```

\subsection*{ VALIDATING THE DATA}
```{r}

# Check number of rows and columns
dim(data)

# Check unique counties 
length(unique(data$county))

# Check duplicates  and missing values
data %>%
  count(county, date) %>%
  filter(n > 1)
colSums(is.na(data))
```
 
\subsection*{Detailed summary statistics}
```{r}
summary(data)
```


```{r}
# Full descriptive statistics table

desc_stats <- data %>%
  select(
    rainfall,
    temperature,
    humidity,
    ndvi,
    population,
    itn,
    irs,
    treatment,
    incidence_per_1000
  ) %>%
  summarise_all(
    list(
      Mean = mean,
      SD = sd,
      Min = min,
      Max = max
    ),
    na.rm = TRUE
  )

View(desc_stats)
```

\section*{HISTOGRAM DISTRIBUTION ON MALARIA INCIDENCE}
```{r}
# Incidence distribution
ggplot(data, aes(x = incidence_per_1000)) +
  geom_histogram(bins = 30, fill = "steelblue") +
  theme_minimal() +
  labs(title = "Distribution of Malaria Incidence")
```
\subsection*{RAINFALL}
```{r}
# Rainfall
ggplot(data, aes(x = rainfall)) +
  geom_histogram(bins = 30, fill = "blue") +
  theme_minimal() +
  labs(title = "Rainfall Distribution")
```

\subsubsection*{Rainfall Distribution per County}
```{r}
ggplot(data, aes(x = county, y = rainfall)) +
  geom_boxplot(fill = "blue") +
  theme_minimal() +
  labs(title = "Rainfall Distribution per County") +
  theme(axis.text.x = element_text(angle = 90))

#saving the plot
ggsave("Rainfall Distribution per County.png", width = 5, height = 4, dpi = 300)
```
\subsection*{TEMPERATURE}
```{r}
# Temperature
ggplot(data, aes(x = temperature)) +
  geom_histogram(bins = 30, fill = "orange") +
  theme_minimal() +
  labs(title = "Temperature Distribution")
```

\subsubsection*{Temperature Distribution per County}

```{r}
#Temperature influences mosquito development rates.
ggplot(data, aes(x = reorder(county, temperature, FUN = median), 
                    y = temperature)) +
  geom_boxplot(fill = "orange", color = "darkred", outlier.alpha = 0.3) +
  coord_flip() +
  labs(title = "Monthly Temperature Variability by County",
       x = "County", 
       y = "Temperature (°C)") +
  theme_minimal()
ggsave("Temperature Distribution per County.png", width = 5, height = 4, dpi = 300)
```
\subsection*{HUMIDITY}
```{r}
# Humidity
ggplot(data, aes(x = humidity)) +
  geom_histogram(bins = 30, fill = "purple") +
  theme_minimal() +
  labs(title = "Humidity Distribution")
```

\subsubsection*{Humidity Distribution per County}


```{r}
# High humidity is a known driver for malaria transmission.
ggplot(data, aes(x = reorder(county, humidity, FUN = median), 
                    y = humidity)) +
  geom_boxplot(
    fill = "lightgreen", 
    color = "darkgreen", 
    outlier.alpha = 0.3
  ) +
  coord_flip() +
  labs(
    title = "Monthly Relative Humidity by County",
    subtitle = "Counties ranked by median humidity levels",
    x = "County",
    y = "Humidity (%)"
  ) +
  theme_minimal()
```
```{r}
data %>%
  select(rainfall, temperature, humidity, ndvi) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot(fill = "skyblue") +
  theme_minimal() +
  labs(title = "Global Comparison of Environmental Variables")
ggsave(" Environmental Comparison variables.png", width = 5, height = 4, dpi = 300)
```
```{r}
data %>%
  select(county, rainfall, temperature, humidity, ndvi) %>%
  pivot_longer(-county, names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = county, y = value)) +
  geom_boxplot(fill = "green") +
  facet_wrap(~variable, scales = "free_y") +
  theme_minimal() +
  labs(title = "Predictor Distribution per County") +
  theme(axis.text.x = element_text(angle = 90))
```
\subsection*{BARPLOT MEAN PER COUNTY}
```{r}
# Convert to long format
data_long <- data %>%
  select(county, rainfall, temperature, humidity, ndvi,
         population, itn, irs, treatment) %>%
  pivot_longer(-county, names_to = "variable", values_to = "value")

# Compute mean values
data_mean <- data_long %>%
  group_by(county, variable) %>%
  summarise(mean_value = mean(value, na.rm = TRUE), .groups = "drop")

# Plot
ggplot(data_mean, aes(x = county, y = mean_value)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  facet_wrap(~variable, scales = "free_y") +
  theme_minimal() +
  labs(title = "Mean Predictor Values per County") +
  theme(axis.text.x = element_text(angle = 90))
```
\subsection*{TEMPORAL TRENDS}
```{r}
selected_counties <- c("Lamu", "Isiolo", "Nairobi", "Kiambu")

data %>%
  filter(county %in% selected_counties) %>%
  # Ensure date is recognized as a Date object
  mutate(date = as.Date(date)) %>% 
  ggplot(aes(x = date, y = incidence_per_1000, group = 1)) + # Added group = 1
  geom_line(colour = "darkgreen") +
  facet_wrap(~ county, scales = "free_y") +
  labs(title = "Temporal Trends by County",
       x = "Year",
       y = "Incidence per 1000") +
  theme_minimal()
ggsave("TEMPORAL TRENDS.png", width = 5, height = 4, dpi = 300)
```
```{r}
#  date conversion first
data$date <- as.Date(data$date)


ggplot(data, aes(x = date, y = incidence_per_1000, group = county, colour = county)) +
  geom_line() +

  scale_x_date(limits = c(as.Date("2016-01-01"), as.Date("2025-12-01")), 
               date_breaks = "1 year", 
               date_labels = "%Y") +
  labs(title = "Malaria Incidence Over Time by County",
       x = "Year",
       y = "Incidence per 1000") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))
ggsave("TEMPORAL TRENDS for  Malaria Incidence Over Time by County.png", width = 5, height = 4, dpi = 300)
```

\subsection*{SEASONALITY OF ANALYSIS}
 
 
```{r}

# Convert the columns explicitly to simple numeric vectors
data$month <- as.numeric(unlist(data$month))
data$incidence_per_1000 <- as.numeric(unlist(data$incidence_per_1000))

ggplot(data, aes(x = factor(month), y = incidence_per_1000)) +
  geom_boxplot(fill = "purple", outlier.colour = "red", outlier.shape = 1) +
  theme_minimal() +
  labs(title = "Seasonal Variation in Malaria Incidence",
       subtitle = "Aggregated monthly distribution (2016-2025)",
       x = "Month",
       y = "Incidence per 1000")
ggsave("SEASONALITY OF ANALYSIS.png", width = 5, height = 4, dpi = 300)
```
 
\subsection*{COUNTY INNCIDENCE COMPARISON}
```{r}
ggplot(data, aes(x = county, y = incidence_per_1000)) +
  geom_boxplot(fill = "red") +
  theme_minimal() +
  labs(title = "Malaria Incidence per County") +
  theme(axis.text.x = element_text(angle = 90))
ggsave("COUNTY INNCIDENCE COMPARISON.png", width = 5, height = 4, dpi = 300)
```
\subsection*{RELATIONSHIP BTW PREDICTORS AND TARGETS}
```{r}
colSums(is.na(data))
```

\subsubsection*{Line plot of annual mean incidence}
```{r}

# Create the annual dataframe from your main data object
annual <- data %>%
  group_by(year) %>%
  summarise(
    Mean_Incidence = mean(incidence_per_1000, na.rm = TRUE),
    SD = sd(incidence_per_1000, na.rm = TRUE)
  )

# Now run your plot code
ggplot(annual, aes(x = year, y = Mean_Incidence)) +
  geom_line(colour = "skyblue", linewidth = 1.2) +
  geom_point(colour = "red", size = 3) +
  geom_errorbar(aes(ymin = Mean_Incidence - SD, 
                    ymax = Mean_Incidence + SD), 
                width = 0.2, colour = "grey") +
  scale_x_continuous(breaks = 2016:2025) +
  labs(
    title   = "Annual Mean Malaria Incidence Rate (2016–2025)",
    subtitle = "Error bars show ± 1 standard deviation across all county-months",
    x = "Year",
    y = "Mean Incidence (cases per 1,000 per month)"
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())
ggsave("Annual Mean Malaria Incidence Rate (2016–2025).png", width = 5, height = 4, dpi = 300)
```
\subsection*{HEATMAP CORRELATION}
```{r}
# Correlation matrix
num_data <- data[, c("rainfall","temperature","humidity",
                     "ndvi","population","itn","irs",
                     "treatment","incidence_per_1000")]

cor_matrix <- cor(num_data, use = "complete.obs")

# Convert to long format - Ensure names match the ggplot call
cor_long <- as.data.frame(as.table(cor_matrix))
names(cor_long) <- c("Var1", "Var2", "Correlation")

# plotting with matching names
ggplot(cor_long, aes(Var1, Var2, fill = Correlation)) + # Use cor_long and Correlation
  geom_tile(color = "black", linewidth = 0.5) +  
  geom_text(aes(label = round(Correlation, 2)), size = 3) + # Use Correlation
  scale_fill_gradient2(low = "lightblue", mid = "white", high = "darkblue", midpoint = 0) +
  labs(title = "Correlation Matrix: Malaria Incidence & Covariates",
       subtitle = "Assessing multicollinearity between environmental and intervention factors",
       x = "",
       y = "") +
  theme_minimal() +
  theme(
    panel.grid.major = element_line(color = "grey70", linewidth = 0.3),  
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.7)  
  )
ggsave("HEATMAP CORRELATION.png", width = 5, height = 4, dpi = 300)
```
Interpretation:
Rainfall shows the strongest positive correlation with malaria incidence $(r=0.54)$, indicating that increased rainfall was associated with higher malaria transmission across counties.
Population density shows a moderate negative correlation with malaria incidence $(r=−0.37)$, suggesting that malaria incidence was generally lower in more densely populated counties within the study period.
Temperature, humidity, NDVI, ITN coverage, IRS, and treatment variables shows weak linear correlations with malaria incidence, indicating that their effects on malaria transmission are likely nonlinear and better captured by machine learning and deep learning models.
```{r}
ggplot(data, aes(x = rainfall, y = incidence_per_1000)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", color = "blue") +
  theme_minimal() +
  labs(title = "Rainfall vs Malaria Incidence")
```




 
 

 







