---
title: "Machine Learning Baseline Models"
author: "Lindah"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    number_sections: true
subtitle: Random Forest | XGBoost | Malaria Risk Prediction Kenya
---

```{r setup, include=FALSE}
options(OutDec = ".", 
        encoding = "UTF-8",
        cli.unicode = FALSE)

knitr::opts_chunk$set(echo=TRUE, 
                      message=FALSE, 
                      warning=FALSE,
                      fig.width=10, 
                      fig.height=6,
                      fig.show = 'hold')
```


#  Loading the Libraries

```{r }
library(dplyr)        
library(readr)        
library(lubridate)    
library(ggplot2)      
library(tidyr)        
library(patchwork)    
library(randomForest) 
library(xgboost)      
library(e1071)        
library(caret)        
library(Metrics)      
library(kableExtra)   

# Setting clean latex 
theme_set(theme_bw())
```


# Loading Data

```{r }
DATA <- "/home/student25/Documents/THESIS/MSC_THESIS_package_corrected - Revised/SPATIO-TEMPORAL THESIS/DATA/"

panel  <- read_csv(paste0(DATA, "kenya_Malaria_panel.csv"))
static <- read_csv(paste0(DATA, "county_static_features.csv"))

response_label <- "Incidence per 1000"

cat("Panel rows:", nrow(panel), " - Counties:", n_distinct(panel$county), "\n")
```


# Cleaning and converting Data
# Convert all columns to correct types THAT IS  DATE AND CONTY
```{r }

panel <- panel %>%
  mutate(
    date               = ymd(date),
    year               = year(date),
    month              = month(date),
    county             = toupper(trimws(county)),
    incidence_per_1000 = as.numeric(incidence_per_1000),
    rainfall           = as.numeric(rainfall),
    temperature        = as.numeric(temperature),
    humidity           = as.numeric(humidity),
    ndvi               = as.numeric(ndvi),
    population         = as.numeric(population),
    itn                = as.numeric(itn),
    irs                = as.numeric(irs),
    treatment          = as.numeric(treatment)
  )

clean_label <- "Incidence per 1000"

cat(" Total missing values:", sum(is.na(panel)), "\n")
```



# Adding Lag Features since mosquito development and breeding depends on environmental effects.
#  Malaria responds to climate 1-3 months later
```{r }
panel <- panel %>%
  arrange(county, date) %>%
  group_by(county) %>%           
  mutate(
    # Rainfall lags establishes Breeding pools and larvae development
    rain_lag1 = lag(rainfall,    1),
    rain_lag2 = lag(rainfall,    2),
    rain_lag3 = lag(rainfall,    3),
    # Temperature lags helps Parasite development 
    temp_lag1 = lag(temperature, 1),
    temp_lag2 = lag(temperature, 2),
    temp_lag3 = lag(temperature, 3),
    

# Humidity lags provide Adult mosquito survival rates
    hum_lag1  = lag(humidity,    1),
    hum_lag2  = lag(humidity,    2),
    hum_lag3  = lag(humidity,    3),
# NDVI lags promotes Vegetation and moisture habitat
    ndvi_lag1 = lag(ndvi,        1),
    ndvi_lag2 = lag(ndvi,        2),
    month_num = as.integer(month(date))
  ) %>%
  ungroup() %>%
  drop_na()  

cat("complete. Rows:", nrow(panel), " - Missing values:", sum(is.na(panel)), "\n")
```



#  Train and Test  data Splitting 

```{r }

train <- panel %>% filter(year <= 2022)
test  <- panel %>% filter(year >= 2023)

# Cleaned output using standard characters for the console
cat("Train rows:", nrow(train), " - From:", format(min(train$date)), "to", format(max(train$date)), "\n")
cat("Test rows:", nrow(test), " - From:", format(min(test$date)), "to", format(max(test$date)), "\n")
```



# Defining Features and Scaling


```{r }
# Predictor columns for standard set for all two models
pred_cols <- c(
  "rainfall", "temperature", "humidity", "ndvi",
  "itn", "irs", "treatment",
  "rain_lag1", "rain_lag2", "rain_lag3",
  "temp_lag1", "temp_lag2", "temp_lag3",
  "hum_lag1",  "hum_lag2",  "hum_lag3",
  "ndvi_lag1", "ndvi_lag2",
  "month_num", "population"
)

# Compute mean and SD from TRAINING data 
tr_mean <- sapply(pred_cols, function(v) mean(train[[v]], na.rm=TRUE))
tr_sd   <- sapply(pred_cols, function(v) sd(train[[v]],   na.rm=TRUE))

# Function to apply scaling
scale_df <- function(df) {
  for (v in pred_cols) df[[v]] <- (df[[v]] - tr_mean[v]) / tr_sd[v]
  return(df)
}

train_s <- scale_df(train)
test_s  <- scale_df(test)

# Feature matrices
X_tr <- as.matrix(train_s[, pred_cols])
X_te <- as.matrix(test_s[, pred_cols])

# Response variable 
Y_tr <- train_s$incidence_per_1000
Y_te <- test_s$incidence_per_1000

# Cleaned output
cat("Features:", length(pred_cols), " - X_tr dimensions:", nrow(X_tr), "by", ncol(X_tr), "\n")
```



# FEATURE CORRELATION HEATMAP


```{r}


library(corrplot)

corr_matrix <- cor(train[, pred_cols])

corrplot(
  corr_matrix,
  method = "color",
  type = "upper",
  tl.cex = 0.7,
  number.cex = 0.5
)
```

# Evaluation Function

```{r }
evaluate <- function(obs, pred, label = "") {
  r   <- rmse(obs, pred)
  m   <- mae(obs, pred)
  r2  <- 1 - sum((obs - pred)^2) / sum((obs - mean(obs))^2)
  
  # 75th percentile threshold
  thr <- quantile(obs, .75, na.rm = TRUE)  
  tp  <- sum(obs >= thr & pred >= thr)
  tn  <- sum(obs <  thr & pred <  thr)
  fp  <- sum(obs <  thr & pred >= thr)
  fn  <- sum(obs >= thr & pred <  thr)
  acc <- (tp + tn) / (tp + tn + fp + fn)
  
  # CLEANED 
  if (nchar(label) > 0) {
    cat(sprintf("[%s] RMSE=%7.5f MAE=%7.5f R2=%6.4f Acc=%6.4f\n",
                label, r, m, r2, acc))
  }
  
  # Data frame output 
  data.frame(Model=label, 
             RMSE=round(r,5), 
             MAE=round(m,5),
             R2=round(r2,4), 
             Accuracy=round(acc,4))
}
```


# Random Forest

```{r rf}
set.seed(123)

# Tune mtry to find best number of features tried at each split
oob <- sapply(c(3, 5, 7, 10), function(m)
  randomForest(X_tr, Y_tr, ntree=100, mtry=m, nodesize=20)$mse[100])

best_m <- c(3, 5, 7, 10)[which.min(oob)]
cat("Best mtry:", best_m, "\n")

# Fit  Random Forest
rf_fit <- randomForest(
  X_tr, Y_tr,
  ntree      = 200,      
  mtry       = best_m,   
  nodesize   = 25,       
  maxnodes   = 30,       
  importance = TRUE      
)

# Predictions
rf_te <- evaluate(Y_te, predict(rf_fit, X_te), "RF-Test")
rf_tr <- evaluate(Y_tr, predict(rf_fit, X_tr), "RF-Train")

# Store predictions 
test_s$rf_pred <- predict(rf_fit, X_te)
```
# IncMSE tells  how much error rises when this feature is randomly shuffled
```{r }
importance(rf_fit) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Feature") %>%
  arrange(desc(`%IncMSE`)) %>%
  head(15) %>%
  ggplot(aes(reorder(Feature, `%IncMSE`), `%IncMSE`, fill = `%IncMSE`)) +
  geom_col(width = 0.7) +
  coord_flip() +
  scale_fill_gradient(low = "lightblue", high = "lightblue4") +
  labs(title = "Random Forest - Variable Importance (Top 15)",
       x = NULL, 
       y = "Percentage Increase in MSE") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```


#  XGBoost

```{r xgb}
# Convert to DMatrix format
dtrain <- xgb.DMatrix(X_tr, label = Y_tr)
dtest  <- xgb.DMatrix(X_te, label = Y_te)

# Model parameters
xgb_p <- list(
  objective        = "reg:squarederror",
  eta              = 0.01,    
  max_depth        = 10,
  subsample        = 0.85,
  colsample_bytree = 0.85,
  min_child_weight = 1,
  gamma            = 0,
  alpha            = 0,
  lambda           = 1
)

# 5-fold cross-validation
set.seed(42)
cv <- xgb.cv(
  params                = xgb_p,
  data                  = dtrain,
  nrounds               = 1000,
  nfold                 = 5,
  metrics               = "rmse",
  early_stopping_rounds = 100,
  verbose               = FALSE,
  print_every_n         = 500
)

# Safe extraction of best iteration
best_nrounds <- if (!is.null(cv$best_iteration) && cv$best_iteration > 0) {
  cv$best_iteration 
} else {
  which.min(cv$evaluation_log$test_rmse_mean)
}

# Clean output for console
cat("Best boosting rounds found:", best_nrounds, "\n")
```

```{r }
# Confirming colomns if they exist
if (is.null(colnames(X_tr))) {
  colnames(X_tr) <- pred_cols
}

# Re-creating dtrain matrix
dtrain <- xgb.DMatrix(X_tr, label = Y_tr)
dtest  <- xgb.DMatrix(X_te, label = Y_te)

# Final model training
xgb_fit <- xgb.train(params = xgb_p, 
                     data = dtrain, 
                     nrounds = best_nrounds, 
                     verbose = 0)

# Predictions and Evaluation
xgb_te_results <- evaluate(Y_te, predict(xgb_fit, dtest),  "XGB-Test")
xgb_tr_results <- evaluate(Y_tr, predict(xgb_fit, dtrain), "XGB-Train")

# Store test predictions
test_s$xgb_pred <- predict(xgb_fit, dtest)
```
# Gain to improvement in accuracy from this feature in splits
```{r }

xgb.importance(feature_names = colnames(X_tr), model = xgb_fit) %>%
  head(15) %>%
  ggplot(aes(reorder(Feature, Gain), Gain, fill = Gain)) +
  geom_col(width = 0.7) +
  coord_flip() +
  scale_fill_gradient(low = "orange", high = "red") +
  labs(title = "XGBoost - Feature Importance (Top 15)", 
       x = NULL, 
       y = "Gain (Contribution to Accuracy)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

Interpretation:By analyzing feature importance, I found that population and rainfall are the most significant predictors of malaria risk in the XGBoost model. In the Random Forest model, rainfall and population also showed the highest percentage increase in Mean Squared Error (MSE) when shuffled, confirming their critical role. Interestingly, climate factors like humidity and NDVI (vegetation index) also contributed significantly, often with lagged effects of 1 to 3 months.



# Final Model Comparison

```{r }
# Combining all test results
final <- do.call(rbind, list(rf_te, xgb_te_results)) %>%
  arrange(RMSE) %>%
  mutate(Rank = row_number())

# Cleaned console 
cat("\n FINAL MODEL COMPARISON \n")
print(final[, c("Rank","Model","RMSE","MAE","R2","Accuracy")], row.names=FALSE)
final %>%
  kbl(caption = "ML Model Performance - Test Period (2023-2025) - Response: Incidence per 1000",
      align   = "clcccc",
      booktabs = TRUE) %>% 
  kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE) %>%
  
  row_spec(1, bold=TRUE, background = "green", color = "white") %>%  
  row_spec(nrow(final), background = "gray!20") %>%
  footnote(general = "Test period: Jan 2023 to Dec 2025 - 47 counties")
```

Interpretation:I compared three machine learning models that is XGBoost and Random Forest to predict malaria incidence. Based on the Root Mean Square Error (RMSE), XGBoost emerged as the top performing model with an RMSE of 0.00863. It also achieved the highest $R^2$ of 0.8199, indicating it explains approximately 82% of the variance in the test data. Random Forest followed in second place ($R^2 = 0.7718$).

# TEMPORAL FORECAST COMPARISON

```{r}


monthly_compare <- test_s %>%
  group_by(date) %>%
  summarise(
    Observed = mean(incidence_per_1000),
    RF = mean(rf_pred),
    XGB = mean(xgb_pred),
  ) %>%
  pivot_longer(-date,
               names_to = "Series",
               values_to = "Incidence")

ggplot(monthly_compare,
       aes(x = date,
           y = Incidence,
           color = Series)) +

  geom_line(linewidth = 1) +

  labs(
    title = "Monthly Malaria Forecast Trends",
    subtitle = "Observed vs Predicted Incidence",
    x = "Date",
    y = "Incidence per 1,000"
  ) +

  theme_bw(base_size = 12)

```


# COUNTY-LEVEL FORECASTS

```{r}


focus_counties <- c(
  "KISUMU",
  "BUSIA",
  "KAKAMEGA",
  "ELGEYO MARAKWET"
)

county_plot <- test_s %>%
  filter(county %in% focus_counties)

ggplot(county_plot,
       aes(x = date)) +

  geom_line(aes(y = incidence_per_1000,
                color = "Observed"),
            linewidth = 1) +

  geom_line(aes(y = xgb_pred,
                color = "XGBoost"),
            linewidth = 1) +

  facet_wrap(~county,
             scales = "free_y") +

  labs(
    title = "County-Level Malaria Forecasts",
    x = "Date",
    y = "Incidence per 1,000",
    color = "Series"
  ) +

  theme_bw(base_size = 11)
```

```{r }
mc <- c("RF-Test" = "forestgreen", 
        "XGB-Test" = "red3" 
    )

mk2 <- function(df, col, ttl, lo = TRUE) {
  df %>%
    ggplot(aes(reorder(Model, if(lo) -.data[[col]] else .data[[col]]),
               .data[[col]], fill = Model)) +
    geom_col(width = 0.7) +
    geom_text(aes(label = round(.data[[col]], 4)),
              hjust = -0.1, size = 3) +
    coord_flip() +
    scale_fill_manual(values = mc) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.25))) +
    # Removed arrows from the title variable
    labs(title = ttl, x = NULL, y = col) +
    theme_minimal(base_size = 10) +
    theme(legend.position = "none",
          # Ensure text is standard 
          text = element_text(family = "sans"))
}

# CLEANING TITLES
p1 <- mk2(final, "RMSE", "RMSE (Lower is better)")
p2 <- mk2(final, "R2",   "R2 (Higher is better)", FALSE)
p3 <- mk2(final, "MAE",  "MAE (Lower is better)")
p4 <- mk2(final, "Accuracy", "Hotspot Accuracy", FALSE)

# Combining using patchwork
(p1 + p2) / (p3 + p4) +
  plot_annotation(
    title    = "Machine Learning Model Comparison",
    subtitle = "Test period: 2023-2025 - Response: Incidence per 1000",
    theme    = theme(plot.title = element_text(face = "bold", size = 13))
  )
ggsave(" Baseline Machine learning metrics.png", width = 5, height = 4, dpi = 300)
```


#  Train vs Test $R^2$ 

```{r }
comparison_data <- data.frame(
  Model   = rep(c("XGBoost","Random Forest"), each = 2),
  Set     = rep(c("Train","Test"), 2),
  R2      = c(xgb_tr_results$R2, xgb_te_results$R2,
              rf_tr$R2,          rf_te$R2
              )
)

ggplot(comparison_data, aes(x = Model, y = R2, fill = Set)) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  geom_text(aes(label = round(R2, 3)), 
            position = position_dodge(0.9), vjust = -0.4, size = 3.5) +
 
  scale_fill_manual(values = c("Train" = "forestgreen", "Test" = "purple")) +
  # CLEANED
  labs(title    = "Train vs Test R2 - Overfitting Diagnostic",
       subtitle = "A large gap between bars indicates potential overfitting",
       x = NULL, 
       y = "R-squared (R2)", 
       fill = "Dataset") +
  theme_bw(base_size = 12)
ggsave("Training vs Testing diagnostic of ML.png", width = 5, height = 4, dpi = 300)
```

Interpretation:I evaluated the gap between training and testing performance to check for overfitting. The XGBoost model showed a very high training $R^{2}$ of 0.9995 compared to its test $R^{2}$ of 0.8202. While this large gap suggests high model complexity, it still maintains the best predictive power on the unseen test set. Random Forest and SVR showed more balanced performance between sets, but with lower overall accuracy on the test data.  
#  Observed vs Predicted — All Models
```{r}
# Join risk_class then reshape to long format for facet plot
static_clean <- static %>%
  mutate(county_join = toupper(trimws(county))) %>%
  select(county_join, risk_class) %>%
  distinct(county_join, .keep_all = TRUE)

plot_ready <- test_s %>%
  mutate(county_join = toupper(trimws(as.character(county)))) %>%
  left_join(static_clean, by = "county_join") %>%
  mutate(risk_class = ifelse(is.na(risk_class), "Unknown", risk_class)) %>%
 
  pivot_longer(cols = c(rf_pred, xgb_pred),
               names_to  = "Model_Type",
               values_to = "Predicted_Val") %>%
  mutate(Model_Name = factor(
    recode(Model_Type,
           "xgb_pred" = "Rank 1: XGBoost",
           "rf_pred"  = "Rank 2: Random Forest"),
    levels = c("Rank 1: XGBoost","Rank 2: Random Forest")
  ))

# Mapping the observed variable to a name without underscores for the plot object
ggplot(plot_ready, aes(x = incidence_per_1000, y = Predicted_Val)) +
  geom_abline(slope=1, intercept=0, linetype="dashed", color="black", linewidth=1) +
  geom_smooth(method="lm", color="purple", se=FALSE, linewidth=1.2) +
  geom_point(aes(fill = risk_class), shape=21, color="black",
             alpha=1, size=1.8, stroke=0.3) +
  facet_wrap(~Model_Name) +
  # Using universal color names
  scale_fill_manual(values = c("Low"      = "forestgreen",
                                "Moderate" = "gold",
                                "High"      = "red3",
                                "Unknown"   = "skyblue4")) +
  labs(title    = "Observed vs Predicted Malaria Incidence",
       subtitle = "Dashed line = perfect prediction - Purple line = model trend",
       x    = "Observed Incidence (per 1,000)",
       y    = "Predicted Incidence (per 1,000)",
       fill = "Risk Class") +
  theme_bw(base_size = 13) +
  theme(legend.position   = "bottom",
        strip.background  = element_rect(fill = "grey90"),
        strip.text        = element_text(face = "bold"))
ggsave( "Observed vs Predicted diagnostic of ML.png", width = 5, height = 4, dpi = 300)
```


\section*{MAPPING AND PREDICTION}


```{r}
library(sf)
library(dplyr)
library(ggplot2)
library(readr)
library(stringr)
library(viridis)
library(tmap)

kenya_map <- st_read("/home/student25/Documents/THESIS/MSC_THESIS_package_corrected - Revised/SPATIO-TEMPORAL THESIS/kenyan-counties/County.shp")

#  LOAD PREDICTIONS


pred <- read.csv("/home/student25/Documents/THESIS/MSC_THESIS_package_corrected - Revised/SPATIO-TEMPORAL THESIS/Rcodes/UPDATED/baseline_predictions.csv")

head(pred)




#  LOAD COUNTY SHAPEFILE
kenya_map <- st_read("/home/student25/Documents/THESIS/MSC_THESIS_package_corrected - Revised/SPATIO-TEMPORAL THESIS/kenyan-counties/County.shp")

print(names(kenya_map))


```
```{r}
county_col <- "COUNTY"
```


# CLEAN COUNTY NAMES

```{r}


# Function to standardize names
clean_names <- function(x){

  x <- toupper(trimws(x))

  # Remove extra spaces
  x <- str_squish(x)

  # FIX SPECIAL COUNTY NAMES
  x <- str_replace_all(x, "-", " ")

  # FIX ELGEYO MARAKWET
  x[x %in% c("ELGEYO-MARAKWET",
             "ELGEYO MARAKWET",
             "KEIYO MARAKWET")] <- "ELGEYO MARAKWET"

  # FIX THARAKA NITHI
  x[x %in% c("THARAKA",
             "THARAKA-NITHI",
             "THARAKA NITHI")] <- "THARAKA NITHI"

  # FIX TRANS NZOIA
  x[x %in% c("TRANSNZOIA",
             "TRANS NZOIA")] <- "TRANS NZOIA"

  # FIX TAITA TAVETA
  x[x %in% c("TAITA-TAVETA",
             "TAITA TAVETA")] <- "TAITA TAVETA"

  return(x)
}

# Apply cleaning
kenya_map$county_join <- clean_names(kenya_map[[county_col]])

pred$county_join <- clean_names(pred$county)

```

#  CHECKING COUNTY MATCHING


```{r}

# Counties in predictions but not shapefile
missing_in_map <- setdiff(unique(pred$county_join),
                          unique(kenya_map$county_join))

print("Missing in shapefile:")
print(missing_in_map)

# Counties in shapefile but not predictions
missing_in_pred <- setdiff(unique(kenya_map$county_join),
                           unique(pred$county_join))

print("Missing in predictions:")
print(missing_in_pred)
```

#  AGGREGATE PREDICTIONS

```{r}


# Average predictions across test period

map_pred <- pred %>%
  group_by(county_join) %>%
  summarise(
    observed = mean(obs, na.rm = TRUE),
    rf_pred = mean(RF, na.rm = TRUE),
    xgb_pred = mean(XGBoost, na.rm = TRUE)
    
  )

head(map_pred)

```

# MERGE WITH SPATIAL DATA

```{r}


kenya_map2 <- kenya_map %>%
  left_join(map_pred, by = "county_join")

```


#  CHECK FOR FAILED MERGES


```{r}

failed <- kenya_map2 %>%
  filter(is.na(xgb_pred)) %>%
  st_drop_geometry()

print("Counties with missing predictions:")
print(failed[, c(county_col, "county_join")])
```

```{r}
options(bitmapType = "cairo")
```
# OBSERVED MALARIA MAP

```{r}


ggplot(kenya_map2) +
  geom_sf(aes(fill = observed),
          color = "black",
          linewidth = 0.2) +
  scale_fill_viridis_c(
    option = "plasma",
    na.value = "grey80"
  ) +
  labs(
    title = "Observed Malaria Incidence in Kenya",
    fill = "Incidence"
  ) +
  theme_minimal()

# SAVE
ggsave("Observed_Malaria_Map.png",
       width = 10,
       height = 8,
       dpi = 300)

```


#  XGBOOST PREDICTED RISK MAP

```{r}


ggplot(kenya_map2) +
  geom_sf(aes(fill = xgb_pred),
          color = "black",
          linewidth = 0.2) +
  scale_fill_viridis_c(
    option = "inferno",
    na.value = "grey80"
  ) +
  labs(
    title = "Predicted Malaria Risk Map (XGBoost)",
    fill = "Predicted Risk"
  ) +
  theme_minimal()

# SAVE
ggsave("XGBoost_Risk_Map.png",
       width = 10,
       height = 8,
       dpi = 300)

```



# SPATIAL PREDICTION ERROR MAP

```{r}


kenya_map2$error_xgb <- kenya_map2$xgb_pred - kenya_map2$observed

ggplot(kenya_map2) +

  geom_sf(aes(fill = error_xgb),
          color = "black",
          linewidth = 0.2) +

 scale_fill_gradient2(
  low = "darkblue",
  mid = "white",
  high = "darkred",
  midpoint = 0,
  limits = c(
    min(kenya_map2$error_xgb),
    max(kenya_map2$error_xgb)
  )
) +

  labs(
    title = "Spatial Prediction Error Map (XGBoost)",
    fill = "Prediction Error"
  ) +

  theme_minimal()
```

Interpretation:
The spatial prediction error map shows the difference between predicted and observed malaria incidence across counties. Counties such as Lamu, Garissa, and parts of eastern Kenya showed negative prediction errors, indicating that the model underestimated malaria incidence in these regions, while several western and central counties showed smaller errors, suggesting better model performance in those areas.
# RANDOM FOREST MAP




```{r}


ggplot(kenya_map2) +
  geom_sf(aes(fill = rf_pred),
          color = "black",
          linewidth = 0.2) +
  scale_fill_viridis_c(
    option = "magma",
    na.value = "grey80"
  ) +
  labs(
    title = "Predicted Malaria Risk Map (Random Forest)",
    fill = "Predicted Risk"
  ) +
  theme_minimal()

ggsave("RF_Risk_Map.png",
       width = 10,
       height = 8,
       dpi = 300)

```

#  CREATE HOTSPOT CLASSES

```{r}

kenya_map2 <- kenya_map2 %>%
  mutate(
    risk_class = case_when(
      xgb_pred <= quantile(xgb_pred, 0.33, na.rm=TRUE) ~ "Low",
      xgb_pred <= quantile(xgb_pred, 0.66, na.rm=TRUE) ~ "Moderate",
      TRUE ~ "High"
    )
  )

table(kenya_map2$risk_class)
```

# HOTSPOT MAP


```{r}




ggplot(kenya_map2) +
  geom_sf(aes(fill = risk_class),
          color = "black",
          linewidth = 0.2) +
  scale_fill_manual(values = c(
    "Low" = "forestgreen",
    "Moderate" = "gold",
    "High" = "red3"
  )) +
  labs(
    title = "Malaria Hotspot Risk Map",
    fill = "Risk Level"
  ) +
  theme_minimal()

ggsave("Hotspot_Risk_Map.png",
       width = 10,
       height = 8,
       dpi = 300)

```

# LABEL COUNTIES ON MAP

```{r}


centroids <- st_centroid(kenya_map2)

ggplot(kenya_map2) +
  geom_sf(aes(fill = xgb_pred),
          color = "red",
          linewidth = 0.2) +

  geom_sf_text(
    data = centroids,
    aes(label = county_join),
    size = 2
  ) +

  scale_fill_viridis_c(option = "inferno") +

  labs(
    title = "Predicted Malaria Risk with County Labels",
    fill = "Risk"
  ) +

  theme_minimal()

ggsave("Labeled_Risk_Map.png",
       width = 12,
       height = 10,
       dpi = 300)
```


#  EXPORTING FINAL SPATIAL DATA
```{r}


st_write(kenya_map2,
         "Kenya_Malaria_Risk_Map.geojson",
         delete_dsn = TRUE)

write.csv(
  st_drop_geometry(kenya_map2),
  "Kenya_Malaria_Risk_Table.csv",
  row.names = FALSE
)

```
```{r}
cat("\n\n")
cat(" MALARIA RISK MAPS GENERATED\n")
cat("\n")

cat("Files saved:\n")
cat("- Observed_Malaria_Map.png\n")
cat("- XGBoost_Risk_Map.png\n")
cat("- RF_Risk_Map.png\n")
cat("- Hotspot_Risk_Map.png\n")
cat("- Labeled_Risk_Map.png\n")
cat("- Kenya_Malaria_Risk_Map.geojson\n")
cat("\n")

```






#  Saving the Outputs

```{r }
prediction_export <- test_s %>% 
  select(county, date, 
         obs     = incidence_per_1000, 
         RF      = rf_pred, 
         XGBoost = xgb_pred
         )

write.csv(prediction_export, "baseline_predictions.csv", row.names = FALSE)

# Save the metrics table (RMSE, MAE, R2, etc.)
write.csv(final, "baseline_metrics.csv", row.names = FALSE)

# Cleaned console output
cat("Success: Results exported to CSV files.\n")
cat("Observations saved:", nrow(prediction_export), "\n")
```

